Reasoning about Connectors in Coq

### 2017-05-26 FACS_Extension/paper_revised0508.tex
paper-updated

\documentclass{llncs}
\usepackage{graphicx}
% used for flow charts
\usepackage{pgf}
\usepackage{tikz}
\usepackage{reotex}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{cite}
\usepackage[normalem]{ulem}
\usepackage{todonotes}
\usepackage{color}
\usepackage{verbatim}
\newcommand{\redt}[1]{\textcolor{red}{#1}}
\newcommand{\liyi}[1]{\textcolor{blue}{#1}}
%\newcommand{\xyz}[1]{\textcolor{blue}{#1}}
\usepackage{url}
\pagestyle{plain}
\begin{document}
\pagestyle{plain}
\mainmatter

@@

\titlerunning{Reasoning about Connectors in Coq}

@@

\authorrunning{X.~Zhang, W.~Hong, Y.~Li and M.~Sun}

\institute{Department of Informatics and LMAM, School of Mathematical Sciences,\\ Peking University\\
\email{\{zhangxiyue,wj.hong,liyi\_math,sunm\}@pku.edu.cn}}

\maketitle
\begin{abstract}
Reo is a channel-based exogenous coordination model in which complex coordinators, called connectors, are compositionally
built out of simpler ones. In this paper, we present a new approach to model connectors in Coq which is a proof assistant based on higher-order logic and $\lambda$-calculus.
The model reflects the original structure of connectors simply and clearly. In our framework, basic connectors (channels) are interpreted as axioms and composition operations are specified as inference rules. Furthermore, connectors are interpreted as logical predicates which describe the relation
between inputs and outputs. With such definitions provided, connector properties, as well as equivalence and refinement relations between different connectors, can be naturally formalized as \emph{goals} in Coq and easily proved using pre-defined \emph{tactics}. Particularly, to fulfil the proof of refinement relation, Z3, which can be used as a component in the context of other tools that require solving logical formulas, works as assistance.
\paragraph*{Keywords:}
Coordination language, Reo, Coq, Reasoning, Z3
\end{abstract}

\section{Introduction}\label{sec:introduction}

@@

In this paper, we aim to provide an approach to formally modeling and reasoning about connectors using Coq.
The basic idea of our
approach is to model the behavior of a connector by representing it as a logical predicate which describes the relation among
the timed data streams on the input and output nodes, and to reason about connectors' properties, as
well as the equivalence and refinement relations
between connectors, by using proof principles and tactics in Coq.
Compared with existing approaches for verifying connectors' properties \cite{BBK+10,KB09,KKV12}, using Coq is especially helpful when
we take infinite behavior into consideration. The coinductive proof principle makes it possible to prove connectors' properties easily while it
is difficult (sometimes impossible) for other approaches (like model checking) because of the huge (or maybe infinite) number of states.

@@

Z3\cite{MouraB08} is an efficient SMT Solver freely available from Microsoft Research. It can be used in various software verification and analysis applications. Z3, acting as a SMT (Satisfiability Modulo Theories) solver, expands to deciding the satisfiability (or dually the validity) of first order formulas with respect to combinations of theories such as: arithmetic, bit-vectors, arrays, and uninterpreted functions.

\section{Basic Definitions}\label{sec:basicdef}

@@

\section{Refinement Checking} \label{sec:refinement}
Modelling and reasoning of Reo connectors in Coq can only provide a proof of one connector is a refinement of the other connector indeed when coming to refinement checking, although it allows us to reason about a relatively wide range of properties. Moreover, refinement relation is an important part of properties, which is used in many applications. Therefore, in order to fulfil the proof of refinement relation, Z3 solver is needed as an assistant tool, applied to the refinement property.
\subsection{Basic Channels and Operators}
Before using Z3 to fulfil the proof of refinement relation, we need to implement the construction of connectors and also define the refinement relation in a way that fits in with Z3 solver. Typically, connector construction starts with the definition of basic channels and composition operators.

Coinciding with the definition of channels in Coq, we dealt with both time and data relation constraints in the basic definition. An auxiliary function \emph{Merge}, different from the composition operation which uses the name \emph{Merger} instead, is used to take the conjunction of every constraint in the constraint lists parameterized by \emph{constraints}.

By means of function \emph{Merge}, basic channels are defined concisely. But the biggest difference is what we feed Z3 solver need to be constraints such that Z3 solver can make decisions and yield results about relations between connectors. For example, the specific behavior constraints for \emph{Sync} channel are classified into two types: data constraints and time constraints , but will all be merged finally.

\begin{verbatim}
def Sync(nodes, bound):
   assert len(nodes) == 2
   constraints = []
   for i in range(bound):
     constraints += [ nodes[0]['data'][i] == nodes[1]['data'][i] ]
     constraints += [ nodes[0]['time'][i] == nodes[1]['time'][i] ]

   return Merge(constraints)
\end{verbatim}

\emph{FIFO1} and \emph{Syncdrain} channel are defined in a similar way. Their corresponding constraint lists consist of constrain statements similar to the statements above.

\emph{Lossy} channel is defined in a recursive way according to its own behavior. And the constraints are in fantastic coincidence with the original definition.
\begin{verbatim}
def Lossy(nodes, bound, idx = 0, num = 0):
   assert len(nodes) == 2
   if bound == num:
     return True
   if bound == idx:
     return True
   constraints_0 = []
   constraints_1 = []
   constraints_0 +=
           [ nodes[0]['data'][idx] != nodes[1]['data'][num]]
   constraints_0 +=
           [ nodes[0]['time'][idx] != nodes[1]['time'][num]]
   constraints_1 +=
           [ nodes[0]['data'][idx] == nodes[1]['data'][num]]
   constraints_1 +=
           [ nodes[0]['time'][idx] == nodes[1]['time'][num]]
   return Or(And(Merge(constraints_0),
             Channel.Lossy(nodes, bound, idx + 1, num)),
             And(Merge(constraints_1),
             Channel.Lossy(nodes, bound, idx + 1, num + 1))
             )
\end{verbatim}

As for composition operators, \emph{Merge} is specially dealt with as a channel \emph{Merger}, defined in a recursive way like \emph{Lossy}. Note that the constraints also conform well to the original definition of \emph{Merge}. Therefore, an additional advantage of defining \emph{Merger} channel in this way is that it can preserve the behavioral pattern to a great extent without many assumptions such as priorities of reading or taking data.
\begin{verbatim}
def Merger(nodes, bound, idx_1 = 0, idx_2 = 0):
   assert len(nodes) == 3
   if bound == idx_1 + idx_2:
     return True
   constraints_1 = []
   constraints_2 = []
   constraints_1 += 
    [ nodes[0]['data'][idx_1] == nodes[2]['data'][idx_1 + idx_2]]
   constraints_1 += 
    [ nodes[0]['time'][idx_1] == nodes[2]['time'][idx_1 + idx_2]]
   constraints_1 += 
    [ nodes[0]['time'][idx_1] <  nodes[1]['time'][idx_2]]
   constraints_2 += 
    [ nodes[1]['data'][idx_2] == nodes[2]['data'][idx_1 + idx_2]]
   constraints_2 += 
    [ nodes[1]['time'][idx_2] == nodes[2]['time'][idx_1 + idx_2]]
   constraints_2 += 
    [ nodes[1]['time'][idx_2] <  nodes[0]['time'][idx_1]]
   return Or(And(Merge(constraints_1), 
             Channel.Merger(nodes, bound, idx_1 + 1, idx_2)),
             And(Merge(constraints_2), 
             Channel.Merger(nodes, bound, idx_1, idx_2 + 1))
             )
\end{verbatim}
\emph{Replicate} and \emph{flow-through} are implicitly implied when we define the connector class and compose channels to construct connectors using the same node names. 
\begin{verbatim}
class Connector:
    def __init__(self):
        self.channels = []

    def connect(self, channel, *nodes):
        self.channels += [(channel, nodes)]
        return self
\end{verbatim}
Simple examples are \emph{c1.connect('Sync', 'A', 'E'), c1.connect('Fifo1', 'E', 'F'), c1.connect('Fifo1', 'E', 'G')} and \emph{c2.connect('Fifo1', 'A', 'D'), c2.connect('Sync', 'D', 'B')}, respectively, in which \emph{c1} and \emph{c2} both belong to connector class.

\subsection{Reasoning about refinement relation}
According to the definition of refinement relation above, we have a formulae to represent properly: $Q \rightarrow P$ if and only if the behavior property of \emph{P} can be derived from theorem \emph{Q}, i.e., the property of connector \emph{Q}. The background theory of the definition of the crucial function  \textbf{\emph{isRefinementOf}} is based on first order logic. Initially, if \emph{Q} is indeed a refinement of \emph{P}, i.e., $Q \rightarrow P$, it can be rephrased as $\neg Q \vee P$. Then, negate it giving $Q \wedge \neg P$. If Z3 solver presents a model satisfying this formula after negation, then the refinement relation is falsified by a counterexample. On the contrary, if there didn't exist a counter example, we may say that the refinement relation, $P\subseteq Q$, holds.

To get some intuition for this, consider the simple case where \emph{sync = Connector(), sync.connect('Sync', 'A', 'B'); fifo1 = Connector(),
fifo1.connect('Fifo1', 'A', 'B')}, and the result of the refinement relation \emph{fifo1.isRefinementOf(sync)} Z3 solver yields is \emph{False}.

In the preceding proof about refinement relation, using specific tactics and strategies requires a good knowledge of Coq. However, Z3 solver is much easier to make use of after the definition of the function \textbf{\emph{isRefinementOf}} and the construction framework of Reo connectors.

The biggest drawback is the failure of proving infinite timed data streams. The efficiency of Z3 solver is well-known. However, in our experiments, this is hardly even the case. Especially, for the connectors consisting of one of the two recursively defined channels, proof can be time-consuming. Hence, tests including recursive definition are often expensive.
\section{Conclusion and Future Work}\label{sec:conclusion}
In this paper, we present a new approach to model and reason about connectors in the Coq system. The model naturally
preserves the original structure of connectors. This also makes the connector description reasonably readable. We
implement the proof of properties for connectors using identified techniques and tactics provided by Coq. Properties are
defined in terms of predicates which provide an appropriate description of the relation among different
timed data streams on the nodes of a connector. All the analysis and verification work are based on the logical framework
where basic channels are viewed as axioms and composition operations are viewed as operators. As we can address
the relation among different timed data streams, we can easily reason about temporal properties as well as equivalence
and refinement relations for connectors.

>>>>

\documentclass[3p,times]{elsarticle}
%\usepackage{ecrc}
\usepackage{graphicx}
% used for flow charts
\usepackage{listings}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{reotex}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

\usepackage[normalem]{ulem}
\usepackage{todonotes}
\usepackage{color}
\usepackage{verbatim}
\newcommand{\redt}[1]{\textcolor{red}{#1}}
\newcommand{\liyi}[1]{\textcolor{blue}{#1}}
%\newcommand{\xyz}[1]{\textcolor{blue}{#1}}

\newtheorem{example}{Example}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]

\usepackage{url}
%\pagestyle{plain}
\begin{document}
\begin{frontmatter}
%\pagestyle{plain}
%\mainmatter

@@

%\titlerunning{Reasoning about Connectors in Coq}

@@

%\authorrunning{X.~Zhang, W.~Hong, Y.~Li and M.~Sun}

\address{Department of Informatics and LMAM, School of Mathematical Sciences,\\ Peking University}
%\email{\{zhangxiyue,wj.hong,liyi\_math,sunm\}@pku.edu.cn}}

%\maketitle
\begin{abstract}
Reo is a channel-based exogenous coordination model in which complex coordinators, called connectors, are compositionally
built out of simpler ones. In this paper, we present a new approach to model connectors in Coq which is a proof assistant based on higher-order logic and $\lambda$-calculus.
The model reflects the original structure of connectors simply and clearly. In our framework, basic connectors (channels) are interpreted as axioms and composition operations are specified as inference rules. Furthermore, connectors are interpreted as logical predicates which describe the relation
between inputs and outputs. With such definitions provided, connector properties, as well as equivalence and refinement relations between different connectors, can be naturally formalized as \emph{goals} in Coq and easily proved using pre-defined \emph{tactics}. Particularly, to fulfil the proof of refinement relation, Z3, which can be used as a component in the context of other tools that require solving logical formulas, works as assistance.
%\paragraph*{Keywords:}

\end{abstract}
\begin{keyword}
Coordination language, Reo, Coq, Reasoning, Z3
\end{keyword}
\end{frontmatter}
\section{Introduction}\label{sec:introduction}

@@

In this paper, we aim to provide an approach to formally modeling and reasoning about connectors using Coq.
The basic idea of our
approach is to model the behavior of a connector by representing it as a logical predicate which describes the relation among
the timed data streams on the input and output nodes, and to reason about connectors' properties, as
well as the equivalence and refinement relations
between connectors, by using proof principles and tactics in Coq.
Compared with existing approaches for verifying connectors' properties \cite{BBK+10,KB09,KKV12}, using Coq is especially helpful when
we take infinite behavior into consideration. The coinductive proof principle makes it possible to prove connectors' properties easily while it
is difficult (sometimes impossible) for other approaches (like model checking) because of the huge (or maybe infinite) number of states. The main drawback of failing to check the complementary refinement relation in Coq is made up by Z3.
While Coq provides a platform to prove properties with a series of tactics and strategies to be considered, Z3 can be used as a component in the context of other tools that require solving logical formulas which is in consistent with the definition of Reo connectors.

@@

Z3\cite{MouraB08} is an efficient SMT Solver and a high performance theorem prover freely available from Microsoft Research. It can be used in various software verification and analysis applications. Acting as a SMT (Satisfiability Modulo Theories) solver, Z3 expands to deciding the satisfiability (or dually the validity) of first order formulas with respect to combinations of theories such as: arithmetic, bit-vectors, arrays, and uninterpreted functions. Given the data and time constraints of connectors, it allows us to decide the satisfiability against properties or refinement relations. Z3 provides bindings for several programming languages. In this paper, we use the Z3 API in Python to construct models and carry out experiments. The following example will provide an intuitive understanding of Z3 solver.
\begin{lstlisting}[frame=single]
x = Int('x')
y = Int('y')

s = Solver()
s.add(x > 10, Or(x + y > 3, x - y < 2))
#Solving constraints in the solver s
print s.check()

#Create a new scope
s.push()
s.add(y < 11)
#Solving updated set of constraints
print s.check()

#Restoring state
s.pop()
#Solving restored set of constraints
print s.check()
\end{lstlisting}
This example contains two constraints at first. From these two constraints, we can see that Z3 supports Boolean operators, such as \emph{And}, \emph{Or}, \emph{Not}, \emph{Implies (implication)}, apart from the operators like $<$, $>$ for comparison.
The command \emph{Solver()} creates a general purpose solver. Constraints can be added using the method \emph{add}. After being added to the solver, the constraints can be said having been asserted in the solver . The method \emph{check()} solves the asserted constraints. Finally, the result is satisfiable (unsatisfiable) if a solution was (not) found, respectively. A solver may fail to solve a system of constraints and \emph{unknown} is returned. Each solver maintains a stack of assertions. The command \emph{push} creates a new scope by saving the current stack size. The command \emph{pop} removes any assertion performed between it and the matching push.
\section{Basic Definitions}\label{sec:basicdef}

@@

\section{Refinement Checking} \label{sec:refinement}
Modelling and reasoning of Reo connectors in Coq can only provide a proof of that one connector is indeed a refinement of the other connector when coming to refinement checking, although it allows us to reason about a relatively wide range of properties. Moreover, the refinement relation is a fairly important part of properties, which is used in many applications. Therefore, in order to fulfil the proof of the refinement relation, Z3 solver is needed as an assistant tool, particularly applied to refinement checking.
\subsection{Basic Channels}
Before using Z3 to fulfil the proof of refinement relation, we need to implement the construction of connectors and also define the refinement checking function in a way that fits in with Z3 solver. Typically, connector construction starts with the definition of basic channels and composition operators. The formalization of basic framework is important and primary, so that we can further formalize our reasoning within this model.

Coinciding with the definition of channels in Coq, we dealt with both time and data relation constraints in the basic definition. An auxiliary function \emph{Conjunction} is used to take the conjunction of every constraint in the constraint lists parameterized by \emph{constraints}.

By means of function \emph{Conjunction}, basic channels are defined concisely and neatly. But the biggest difference is that what we feed Z3 solver need to be constraints such that Z3 solver can make decisions and yield results about refinement relations between connectors. For example, the specific behavior constraints for \emph{Sync} channel are classified into two types: data constraints and time constraints. The description and the definition of \emph{Sync} in Coq have already clearly expressed the behavior pattern it needs to follow: each item in data and time streams of output specified as ``node[1]'' is equivalent to the data and time streams of input specified as ``node[0]'', which are exactly data and time constraints. As they are both requirements we need to satisfy no matter data or time is related, each constraint in the list will all be merged finally.
\lstset{language=Python}
\begin{lstlisting}[frame=single]
def Sync(nodes, bound):
   assert len(nodes) == 2
   constraints = []
   for i in range(bound):
     constraints += [ nodes[0]['data'][i] == nodes[1]['data'][i] ]
     constraints += [ nodes[0]['time'][i] == nodes[1]['time'][i] ]
   return Merge(constraints)
\end{lstlisting}

\emph{Syncdrain} and \emph{FIFO1} channel are defined in a similar way. On the basis of the behavior of \emph{Syncdrain} channel, there are only time related constraints to be needed, i.e., equivalence of each item in two time streams of the two inputs. Regarding \emph{FIFO1} channels, constraints related data is the same as \emph{Sync} channels. But there are more time related constraints. As the buffer capacity is one, the relation of each item in the input time stream and output time stream need to be carefully dealt with, especially that the next input should be later than the output at present. At last, their corresponding constraint lists consist of constraint statements similar to the statements above.
\begin{lstlisting}[frame=single]
def SyncDrain(nodes, bound):
   assert len(nodes) == 2
   constraints = []
   for i in range(bound):
       constraints +=
            [nodes[0]['time'][i] == nodes[1]['time'][i]]
   return Merge(constraints)
\end{lstlisting}
\begin{lstlisting}[frame=single]
def Fifo1(nodes, bound):
   assert len(nodes) == 2
   constraints = []
   for i in range(bound):
      constraints +=
                  [ nodes[0]['data'][i] == nodes[1]['data'][i] ]
      constraints +=
                  [ nodes[0]['time'][i] <  nodes[1]['time'][i] ]
      if i:
          constraints +=
                  [ nodes[0]['time'][i] > nodes[1]['time'][i-1] ]

   return Merge(constraints)
\end{lstlisting}

In accordance with the behavior of \emph{Lossy} channel, each item of the input stream may or may not be lost. If so, then the output gets nothing. If not, it behaves exactly like \emph{Sync} channel, so in this case the data and time related constraints are identical with those in the definition of \emph{Sync} channel. The biggest difference is that \emph{Lossy} channel should be defined in a recursive way according to its own distinctive behavior. Note that the constraints are added in a recursive way which is in great coincidence with the original definition.
\begin{lstlisting}[frame=single]
def Lossy(nodes, bound, idx = 0, num = 0):
   assert len(nodes) == 2
   if bound == num:
     return True
   if bound == idx:
     return True
   constraints_0 = []
   constraints_1 = []
   constraints_0 +=
           [ nodes[0]['data'][idx] != nodes[1]['data'][num]]
   constraints_0 +=
           [ nodes[0]['time'][idx] != nodes[1]['time'][num]]
   constraints_1 +=
           [ nodes[0]['data'][idx] == nodes[1]['data'][num]]
   constraints_1 +=
           [ nodes[0]['time'][idx] == nodes[1]['time'][num]]
   return Or(And(Merge(constraints_0),
             Channel.Lossy(nodes, bound, idx + 1, num)),
             And(Merge(constraints_1),
             Channel.Lossy(nodes, bound, idx + 1, num + 1))
             )
\end{lstlisting}
\subsection{Composition Operators}
While the basic channels are already well defined as the basis of the whole construction framework, the composition operators also need to be specified for the large-scale connector construction just like cement for bricks.

Among the three types of composition operators, the most important and difficult one is \emph{Merge}, which is specially dealt with as a channel \emph{Merger} in our model. Although the composition operator \emph{Merge} is treated as a channel \emph{Merger}, it also play the same part in the construction of a connector.  Plus the rest of composition operators: \emph{replicate} and \emph{flow-through}, they are enough for the whole construction for any connector, which can serve as same as the original composition operations. For each item of output stream, there is a nondeterministic choice between the two input nodes. Hence, the \emph{Merger} channel also need to be defined in a recursive way like \emph{Lossy} channel. Note that the constraints also conform well to the original definition of \emph{Merge}. Concluded from the two recursive definitions, an additional advantage of defining \emph{Lossy} and \emph{Merger} channel in this way is that it can preserve the behavioral pattern to a great extent without many assumptions such as priorities of reading or taking data.
\begin{lstlisting}[frame=single]
def Merger(nodes, bound, idx_1 = 0, idx_2 = 0):
   assert len(nodes) == 3
   if bound == idx_1 + idx_2:
     return True
   constraints_1 = []
   constraints_2 = []
   constraints_1 +=
    [ nodes[0]['data'][idx_1] == nodes[2]['data'][idx_1 + idx_2]]
   constraints_1 +=
    [ nodes[0]['time'][idx_1] == nodes[2]['time'][idx_1 + idx_2]]
   constraints_1 +=
    [ nodes[0]['time'][idx_1] <  nodes[1]['time'][idx_2]]
   constraints_2 +=
    [ nodes[1]['data'][idx_2] == nodes[2]['data'][idx_1 + idx_2]]
   constraints_2 +=
    [ nodes[1]['time'][idx_2] == nodes[2]['time'][idx_1 + idx_2]]
   constraints_2 +=
    [ nodes[1]['time'][idx_2] <  nodes[0]['time'][idx_1]]
   return Or(And(Merge(constraints_1),
             Channel.Merger(nodes, bound, idx_1 + 1, idx_2)),
             And(Merge(constraints_2),
             Channel.Merger(nodes, bound, idx_1, idx_2 + 1))
             )
\end{lstlisting}
\emph{Replicate} and \emph{flow-through} operators are implicitly achieved when we compose channels to construct connectors using the same node names. A simple example for the implementation of \emph{replicate} is \emph{c1.connect(`Sync', `A', `E'), c1.connect(`Fifo1', `E', `F'), c1.connect(`Fifo1', `E', `G')} where message from node `E' is copied to send to nodes `F' and `G' simultaneously. As for \emph{flow-through} operator, a simple example is \emph{c2.connect(`Fifo1', `A', `D'), c2.connect(`Sync', `D', `B')} where the timed data flows from node `A' to node `B' through the mixed node `D'. In these two simple examples, \emph{c1} and \emph{c2} both belong to connector class which will be elaborated in detail in next module.
\subsection{Reasoning about refinement relation}
With basic channels and composition operators well formulated above, we are left with the task of first constructing Reo connectors. In this model, connector is defined as a new class, which provides a method of constructing complex connectors out of channels and operators.
The statements inside a class definition are function definitions. We have seen the function `connect' used in the two simple examples of \emph{replicate} and \emph{flow-through}.
\begin{lstlisting}[frame=single]
class Connector:
    def __init__(self):
        self.channels = []

    def connect(self, channel, *nodes):
        self.channels += [(channel, nodes)]
        return self
\end{lstlisting}
Another function \textbf{\emph{isRefinementOf}} is the most crucial one in refinement checking. The result of the function \textbf{\emph{isRefinementOf}} is Boolean, i.e., a connector either is or not a refinement of another connector. According to the definition of refinement relation in section \ref{sec:verification}, we have a formula to represent it properly: $Q \rightarrow P$ if and only if the behavior property of \emph{P} can be derived from the property of connector \emph{Q}, i.e., $\emph{P} \sqsubseteq \emph{Q}$. The background theory of the definition of function  \textbf{\emph{isRefinementOf}} is based on first order logic. Initially, if \emph{Q} is indeed a refinement of \emph{P}, i.e., $Q \rightarrow P$, it can be rephrased as $\neg Q \vee P$. Then, negate it giving $Q \wedge \neg P$. If Z3 solver presents a model satisfying this formula after negation, then the refinement relation is falsified by a counterexample. On the contrary, if there didn't exist a counter example, we may say that the refinement relation, $P\subseteq Q$, holds within the given bound. The formal definition of this function \textbf{\emph{isRefinementOf}} is given as an algorithm here.
\begin{algorithm}[t]
    \caption{Function: P.isRefinementOf (Q, bound)}
    \begin{algorithmic}[1]
        \REQUIRE Both P and Q are connector.
        \ENSURE A boolean: \emph{True} or \emph{False with a counterexample}
        \FOR{channel \emph{t} $\leftarrow$ \emph{P}}
        	\FOR{node \emph{n} $\leftarrow$ \emph{t}}
				\IF{\emph{n} is a new node}
					\STATE i.generate variables like 					
						\newline \{`time':[n\_t\_0, n\_t\_1,...,n\_t\_(bound-1)],				
						\newline `date':[n\_d\_0, n\_d\_1,...,n\_d\_(bound-1)]\}
						\newline
						\newline ii.add time constraints like
						\newline ``n\_t\_0 $\geq$ 0  and  n\_t\_i $<$ n\_t\_i+1''
						\newline into solver
				\ENDIF
			\ENDFOR
		\STATE Add the constraints between different nodes of the same channel into solver according to the property of channels.
        \ENDFOR
        \STATE foralls = []      \% to store new variable
        \STATE absGlobalConstr = None \% to store all constraints under Q
        \STATE absTimeConstr = None \% to store all timed constraints under Q
        \FOR{channel \emph{t} $\leftarrow$ \emph{Q}}
        	\FOR{node \emph{n} $\leftarrow$ \emph{t}}
				\IF{\emph{n} is a new node}
					\STATE i.generate variables like 					
						\newline \{`time':[n\_t\_0, n\_t\_1,...,n\_t\_(bound-1)],				
						\newline `date':[n\_d\_0, n\_d\_1,...,n\_d\_(bound-1)]\}
						\newline
						\newline ii.add \emph{n} into forall
						\newline
						\newline iii.add time constraints like
						\newline ``n\_t\_0 $\geq$ 0  and  n\_t\_i $<$ n\_t\_i+1''
						\newline into absTimeConstr
				\ENDIF
			\STATE add the constraints between different nodes of the same channel into absGlobalConstr according to the property of channels.
			\ENDFOR
		\IF{absTimeConstr is not empty}
			\STATE $absGlobalConstr = \neg (absTimeConstr \cap absGlobalConstr)  $
		\ELSE
			\STATE $absGlobalConstr = \neg absGlobalConstr$
		\ENDIF
		\IF{forall is not empty}
			\STATE apply \emph{ForAll} to absGlobalConstr
			\newline add those constraints into solver
	    \ELSE
	    	\STATE add absGlobalConstr into solver
		\ENDIF	
        \ENDFOR
    \STATE Check.
    \end{algorithmic}
\end{algorithm}

\begin{example}
To get some intuition for this, consider a simple case where \emph{sync = Connector(), sync.connect(`Sync', `A', `B'); fifo1 = Connector(),
fifo1.connect(`Fifo1', `A', `B')}. Two simple connectors which is actually \emph{Sync(A,B)} and \emph{Fifo1(A,B)} have been constructed. Note that a \emph{FIFO1} channel is not a refinement of a \emph{Sync} channel. After invoking the function \textbf{\emph{isRefinementOf}}, we can obtain the result of the refinement relation \emph{fifo1.isRefinementOf(sync)}. The result and the counterexample Z3 solver yields are presented as follows.
\begin{verbatim}
False
A_d_0 = 0, A_d_1 = 0, A_d_2 = 0, A_d_3 = 0, A_d_4 = 0,
A_d_5 = 0, A_d_6 = 0, A_d_7 = 0, A_d_8 = 0, A_d_9 = 0,
A_t_0 = 0, A_t_1 = 2, A_t_2 = 4, A_t_3 = 6, A_t_4 = 8, 
A_t_5 = 10,A_t_6 = 12,A_t_7 = 14,A_t_8 = 16,A_t_9 = 18,
B_d_0 = 0, B_d_1 = 0, B_d_2 = 0, B_d_3 = 0, B_d_4 = 0,
B_d_5 = 0, B_d_6 = 0, B_d_7 = 0, B_d_8 = 0, B_d_9 = 0,
B_t_0 = 1, B_t_1 = 3, B_t_2 = 5, B_t_3 = 7, B_t_4 = 9,
B_t_5 = 11,B_t_6 = 13,B_t_7 = 15,B_t_8 = 17,B_t_9 = 19.
\end{verbatim}
Note that the counterexample is easy to understand. There exists a timed data stream whose time satisfies the constraints of \emph{FIFO1} channel, but don't satisfy the time constraints of \emph{Sync} channel.
\end{example}
\begin{example}
Another simple example is also to seek the refinement relation between two basic channels. The construction is similar to the above example: \emph{sync = Connector(), sync.connect(`Sync', `A', `B'); lossy = Connector(), lossy.connect(`Lossy', `A', `B')}. We know that \emph{Sync} channel is a refinement of \emph{Lossy} channel. When \emph{Lossy} channel behaves well enough to lose any data items, then it behaves exactly like \emph{Sync} channel. On the contrary, \emph{Lossy} channel is not a refinement of \emph{Sync} channel which is consistent with the result \emph{False} given by the Z3 solver. The counterexample is also very intelligible. Note that the first four timed data items were lost with the last timed data item being transferred successfully, which is consistent with the constraints of \emph{Lossy} channel and obviously not in concordance with the constraints of \emph{Sync} channel.
\begin{verbatim}
False
A_d_0 = 2, A_d_1 = 0, A_d_2 = 2, A_d_3 = 3, A_d_4 = 1,
A_t_0 = 0, A_t_1 = 1, A_t_2 = 2, A_t_3 = 3, A_t_4 = 4,
B_d_0 = 1, 
B_t_0 = 4.
\end{verbatim}
\end{example}

Before demonstrating complementary refinement checking between large-scale connectors, we will present some results of bounded refinement and equivalence relation checking in the next two examples.
\begin{example}
\label{ex:equivalence1}
In this example, we consider about the two connectors in Fig.\ref{refine} again. In  Example \ref{ex:refine}, we have seen the refinement relation checking between the two connectors in Coq. However, tactics and strategies can be too professional to comprehend or to tackle with. Although Z3 can only provide a bounded refinement relation checking for these two connectors, it will be much easier to check the refinement relation as we don't need to master concrete theorem proving tactics. All we need to do is to constructing the two connectors in the way accepted by Z3 solver and use the function \emph{isRefinementOf}.
\begin{lstlisting}[frame=single]
c1 = Connector()
c1.connect('Sync', 'A', 'M')
c1.connect('Fifo1', 'M', 'B')
c1.connect('Fifo1', 'M', 'C')

c2 = Connector()
c2.connect('Fifo1', 'A', 'N')
c2.connect('Sync', 'N', 'B')
c2.connect('Sync', 'N', 'C')
\end{lstlisting}
Notice that the \emph{replicate} operator is applied twice in the construction process. Assume the bound is ten and then carry out the function \emph{c2.isRefinementOf(c1, 10)}. Finally, we can get the result \emph{True} which means that \emph{Connector Q} in Fig.\ref{refine} is indeed a refinement of \emph{Connector P}. We can further extend the bound and the time complexity increases linearly as no recursively defined channels are involved.
\end{example}
\begin{example}
This example checks the equivalence relation between connectors in Fig.\ref{refine1}. Construct the two connectors first similar to the example \ref{ex:equivalence1} and invoke the function \emph{isRefinementOf} in two directions. Then we can get the equivalence result.
When giving the bound ten, the results of two directions are both \emph{True}. Hence, we can conclude that the two connectors are in equivalence relation.
\end{example}

\begin{example}
In this example, we will take the well-known Reo connector \emph{Exclusive Router} into consideration. This connector and its variant are shown as follows. The \emph{variant router} is defined like \emph{router} with one major difference: one of the two \emph{Lossy} channels is replaced with a \emph{Sync} channel. Therefore, the behavior of the variant router is actually not exclusive and all data items will be transported to node \emph{E} rather than transmitted to node \emph{E} and \emph{F} exclusively in the original router. Notice that the variant router is a refinement of the original router and conversely not. Construct these two connectors first and set the bound to ten. Invoke the function \emph{isRefinementOf} in two directions, i.e., \emph{c2.isRefinementOf(c1, 10), c1.isRefinementOf(c2, 10)} where \emph{c2} is the variant router. We can find the results Z3 solver yields is \emph{True} and \emph{False} respectively, which is consistent with our intuition.
\begin{figure}
\centering
\begin{tikzpicture}
    % part 1. router
    \node (note1) at (6.1, -1.6) {\small (a) Router};
    \node (note2) at (12.5, -1.6) {\small (b) Variant Router};
    \ionode{(in)}{(4.5, 0)}{node[left] {A}}
    \ionode{(outup)}{(7.5, 0.8)}{node[right] {E}}
    \ionode{(outdown)}{(7.5, -0.8)}{node[right] {F}}
    \mixednode{(mid)}{(6, 0)}{node[right] {C}}
    \mixednode{(wayup)}{(6, 0.8)}{node[above] {B}}
    \mixednode{(waydown)}{(6, -0.8)}{node[below] {D}}
    \lossysync{(in)}{(wayup)}{}
    \lossysync{(in)}{(waydown)}{}
    \syncdrain{(in)}{(mid)}{}
    \sync{(wayup)}{(mid)}{}
    \sync{(waydown)}{(mid)}{}
    \sync{(wayup)}{(outup)}{}
    \sync{(waydown)}{(outdown)}{}
    \ionode{(pin)}{(11, 0)}{node[left] {A}}
    \ionode{(poutup)}{(14, 0.8)}{node[right] {E}}
    \ionode{(poutdown)}{(14, -0.8)}{node[right] {F}}
    \mixednode{(pmid)}{(12.5, 0)}{node[right] {C}}
    \mixednode{(pwayup)}{(12.5, 0.8)}{node[above right] {B}}
    %\mixednode{(pmap2)}{(6.5, 1.8)}{node [right] {B'}}
    \mixednode{(pwaydown)}{(12.5, -0.8)}{node[below] {D}}
%    \syncdrain{(pmap2)}{(pwayup)}{}
    \sync{(pin)}{(pwayup)}{}
    \lossysync{(pin)}{(pwaydown)}{}
    \syncdrain{(pin)}{(pmid)}{}
    \sync{(pwayup)}{(pmid)}{}
    \sync{(pwaydown)}{(pmid)}{}
    \sync{(pwayup)}{(poutup)}{}
    \sync{(pwaydown)}{(poutdown)}{}
\end{tikzpicture}
\caption{Example of connector complementary refinement }
\end{figure}

\end{example}
The following large-scale example is aimed at complementary refinement checking.
\subsection{Z3 vs Coq}
Having seen the examples proved by Coq or Z3 solver, we can summarize their features and make a comparison between them.
In the preceding proof about refinement relation or other properties in Coq, the main feature is that users need to master the tactics and strategies well to give a proof of the properties. But there is no need to worry the time consumed in the proof. It will be efficient no matter how large the connectors are as long as users have choose proper tactics. Besides, the connectors whose properties have been proved can be regarded as a theorem when proving properties of much larger ones. Recall that the reason of introducing Z3 solver is that complementary refinement checking, i.e., one connector is not  a refinement of the other one, can't be provided by Coq, let alone provide counterexamples or witnesses.
